Do we need detailed demographic data to forecast plant population responses to climate change?

### 2014-12-11 manuscript/manuscript.tex
adding markdown ms file

---
layout: 12pt,review
linenumbers: true
title: "Do we need detailed demographic data to forecast plant population responses to climate change?"
author: 
  - name: Andrew T. Tredennick
    affiliation: usu
    email: atredenn(at)gmail.com
    footnote: Corresponding author
  - name: Peter B. Adler
    affiliation: usu
address: 
  - code: usu
    address: | 
      Department of Wildland Resources and the Ecology Center, 
      Utah State University,
      Logan, UT 84322, USA
abstract: |
  Forecasts of populations under future climate change are constrained by the spatial and temporal scales at which we can gather data. For plants, ecologists often require individual level data to build population models; data that is difficult and expensive to collect at spatial scales greater than one square meter. Models built using such data are difficult to scale-up to spatial scales relevant to land management decisions and policy intervention. A possible route forward are density-structured models based on coarse-scale census data, which are much easier to collect over large spatial extents. However, to use coarse-scale data in a climate change context requires the assumption that the climate signals in population dynamics identified using individual-level data can also be identified when individual-level data are averaged over. Here we use a longterm dataset to model population dynamics using individual-level and plot-level data.

keywords:
  - ecological forecasts
  - population models
  - integral projection model
  - climate change
  - grassland

bibliography: components/references.bib
csl: components/ecology.csl
documentclass: components/elsarticle

## rmarkdown render options
output:
  pdf_document:
    fig_caption: true
    template: components/elsarticle.latex
    keep_tex: true

---

```{r caching, include=FALSE}
library("methods")
library("knitr")
basename <- "manuscript"
opts_chunk$set(fig.path = paste("components/figure/", basename, "-", sep=""),
               cache.path = paste("components/cache/", basename, "/", sep=""))
opts_chunk$set(cache = 2)
opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, 
               comment = NA, verbose = TRUE, echo=FALSE)

# PDF-based figures
opts_chunk$set(dev='pdf')
```

```{r plot-options, message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

opts_chunk$set(dev='pdf', dev.args=c(version="1.3"))

```

Introduction
============
Producing skillfull forecasts at spatial scales relevant to global environmental changes that play out at the landscape and regional scale has proven challenging for population ecologists. The major hurdle is that most population models, at least for plant species, are built using data from small, localized plots because parameterizing vital rate regressions requires tracking the fates of individuals. These models are difficult to scale up from the micro to meso-scales because the fitted parameters do not fully represent the spatial variation present at spatial scales beyond that at which the data are collected.

@@

Materials and Methods
=====================
### Study site and data
Our demographic data comes from the Fort Keogh Livestock and Range Research Laboratory in eastern Montana's northern mixed prairie near Miles City, Montana, USA (46 deg. 19' N, 105 deg 48' W). The dataset is freely available on Ecological Archives (CITE), and interested readers should refer to the metadata therin for a complete description. The site is about 800 m above sea level and mean annual precipitation (1878-2009) is 334 mm, with most annual precipitaion falling from April through September (76). The site is grass dominated and, for the purposes of our study, we focus on the four most abundant graminoid species: \emph{Bouteloua gracilis}, \emph{Hesperostipa comata}, \emph{Pascopyrum smithii}, and \emph{Poa secunda}.

@@

### Stastical models of vital rates
The first step in building our single-species population models was to fit statistical models of vital rates at both levels of inference: individual-level and quadrat-level. Here we describe the general statistical approach at the individual-level, but the same approach applies at the quadrat-level. We modeled survival and growth as functions of genet size (quadrat cover in the quadrat-level models) and climate covariates (described in more detail below). We maintained a consistent random effects structure for both models that included three terms: (1) a random year effect on the intercept, (2) a random year effect on the coefficient for plant cover (either individual or quadrat level), and (3) a random effect of group (see Data set description) on the intercept. Vital rates for each species are modeled separately.

@@

where $x$ is the log of genet size (quadrat areal cover), $\beta_{0,t}$ is a year-specific intercept, $\beta_{s,t}$ is the year-specific slope parameter for size, $\beta_{Q}$ is the random effect of quadrat group location, $\theta$ is a matrix of $i$ climate effects over $t$ years, $\beta_{c,i}$ is the fixed parameter for the effect of the $i$th climate covariate, and $\varepsilon_{t}$ is the error term. At the quadrat level we did not fit random year effects for survival because those parameters did not reach convergence (see "Computing").

@@

### Including climate covariates (and avoiding model selection)
For both the individual-level IPM (IPM) and the quadrat-based IBM (QBM) we included four climate covariates in each vital rate regression: fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year. We included climate covariates as additive effects. We did not consider interactions among climate covariates or among climate covariates and plant size or proportional cover. Climate effects ($\beta_{c}$s) were modeled hierarchically in each vital rate regression so that each species specific climate effect is drawn from an interspecific climate effect distribution. This was necessary to achieve convergence at the quadrat level and we used the same approach at the individual level for uniformity.

### Model fitting
All of our analyses (model fitting and simulating) were conducted in Program R. We used the MCMC sampler in JAGS to estimate the posterior distributions of model parameters. We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 50,000 iterations, after a 50,000 iteration burn in. We assessed convergence visually and using the Gelman diagnostic in the R package `coda.' Scale reduction factors for all parameters were less than 1.02, indicating convergence. For the purposes of introducing stochasticity in our population models, we saved the final 1,000 iterations from each chain for all parameters to be used as randomly drawn values during population simulation.

### Population models
With the posterior distribution of the vital rate statistical models in hand, it is straightforward to simulate the population models.

Results
=======
We assessed the statistical importance of the climate covariates included the final vital rate regressions by comparing the residual deviance of models with climate covariates and temporal random effects, climate covariates only, and temporal random effects only. When a model includes climate covariates, this comparison shows the relative contribution of the climate covariates in explaining the total interannual variability (Adler et al. 2012).

```{r figure_1, dependson="plot-options", fig.cap="The proportion of interannual variability in  vital rates explained by the climate covariates. The contribution for growth is defined as: (Climate model - Constant Model)/(Full model - Constant model). The contribution for survival and colonization, where we could not estimate a full model with year random effects at the quadrat level, is defined as: (Constant Model - Climate Model)/Constant Model.", fig.width=7, fig.height=3, cache=FALSE, results='hide'}
deviance <- data.frame(VitalRate = c("Growth", "Survival", "Colonization"),
                       ConstantModel = rep(NA,3),
                       ClimateModel = rep(NA,3),
                       FullModel = rep(NA,3),
                       ClimateContribution = rep(NA,3))
deviance$FullModel[1] <- round(read.csv("../montana/quadBM/vitalRateRegressions/growth/growthDeviance.csv")[,2],2)
deviance$ClimateModel[1] <- round(read.csv("../montana/quadBM/vitalRateRegressions/growth/growthDevianceCLIMATE.csv")[,2],2)
deviance$ConstantModel[1] <- round(read.csv("../montana/quadBM/vitalRateRegressions/growth/growthDevianceCONSTANT.csv")[,2],2)
deviance$ClimateContribution[1] <- with(deviance[1,], (ClimateModel-ConstantModel)/(FullModel-ConstantModel))
deviance$ClimateModel[2] <- round(read.csv("../montana/quadBM/vitalRateRegressions/survival/survivalDeviance.csv")[,2],2)
deviance$ConstantModel[2] <- round(read.csv("../montana/quadBM/vitalRateRegressions/survival/survivalDevianceCONSTANT.csv")[,2],2)
deviance$ClimateModel[3] <- round(read.csv("../montana/quadBM/vitalRateRegressions/colonization/colonizationDeviance.csv")[,2],2)
deviance$ConstantModel[3] <- round(read.csv("../montana/quadBM/vitalRateRegressions/colonization/colonizationDevianceCONSTANT.csv")[,2],2)
deviance$ClimateContribution[2:3] <- with(deviance[2:3,], (ConstantModel-ClimateModel)/(ConstantModel))

colnames(deviance) <- c("Vital Rate", "Constant model", "Climate model", "Full model", "Contribution of climate covariates")
library(ggplot2)
devGQ <- data.frame(VitalRate = c("Growth", "Survival", "Colonization"),
                       ConstantModel = rep(NA,3),
                       ClimateModel = rep(NA,3),
                       FullModel = rep(NA,3),
                       ClimateContribution = rep(NA,3))
devGQ$FullModel[1] <- round(read.csv("../montana/ipm/vitalRateRegs/growth/growthDeviance.csv")[,2],2)
devGQ$ClimateModel[1] <- round(read.csv("../montana/ipm/vitalRateRegs/growth/growthDevianceCLIMATE.csv")[,2],2)
devGQ$ConstantModel[1] <- round(read.csv("../montana/ipm/vitalRateRegs/growth/growthDevianceCONSTANT.csv")[,2],2)
devGQ$FullModel[2] <- round(read.csv("../montana/ipm/vitalRateRegs/survival/survivalDeviance.csv")[,2],2)
devGQ$ClimateModel[2] <- round(read.csv("../montana/ipm/vitalRateRegs/survival/survivalDevianceCLIMATE.csv")[,2],2)
devGQ$ConstantModel[2] <- round(read.csv("../montana/ipm/vitalRateRegs/survival/survivalDevianceCONSTANT.csv")[,2],2)
devGQ$FullModel[3] <- round(read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentDeviance.csv")[,2],2)
devGQ$ClimateModel[3] <- round(read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentDevianceCLIMATE.csv")[,2],2)
devGQ$ConstantModel[3] <- round(read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentDevianceCONSTANT.csv")[,2],2)
devGQ$ClimateContribution[1] <- with(devGQ[1,], (ClimateModel-ConstantModel)/(FullModel-ConstantModel))
devGQ$ClimateContribution[2] <- with(devGQ[2,], (ConstantModel-ClimateModel)/(ConstantModel))
devGQ$ClimateContribution[3] <- with(devGQ[3,], (ConstantModel-ClimateModel)/(ConstantModel))

devM <- devGQ[,c(1,5)]
devM$Level <- rep("Individual", 3)
tmpD <- deviance[,c(1,5)]
tmpD$Level <- rep("Quadrat", 3)
colnames(tmpD)[2] <- "value"
colnames(tmpD)[1] <- "VitalRate"
colnames(devM)[2] <- "value"
plotD <- rbind(tmpD, devM)
gClimCont1 <- ggplot(data=subset(plotD, Level=="Individual"), aes(x=VitalRate, y=value))+
  geom_bar(stat="identity", position="dodge", fill="grey30")+
  ylab("Contribution of Climate Effects")+
  xlab("Vital Rate")+
  ggtitle("Individual level")+
  scale_y_continuous(limits=c(0, 0.5))+
  scale_x_discrete(labels=c("Recruitment", "Growth", "Survival"))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
gClimCont2 <- ggplot(data=subset(plotD, Level=="Quadrat"), aes(x=VitalRate, y=value))+
  geom_bar(stat="identity", position="dodge", fill="grey30")+
  ylab("Contribution of Climate Effects")+
  xlab("Vital Rate")+
  ggtitle("Quadrat level")+
  scale_y_continuous(limits=c(0, 0.5))+
  scale_x_discrete(labels=c("Colonization", "Cover change", "Extirpation"))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
library(gridExtra)
gClim <- grid.arrange(gClimCont1, gClimCont2, ncol=2, nrow=1)
gClim

```

>>>>

\documentclass[author-year, 12pt,review]{components/elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}
\usepackage{lineno} % add 
  \linenumbers % turns line numbering on 
\bibliographystyle{elsarticle-harv}
\biboptions{sort&compress} % For natbib
\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%% Redefines the elsarticle footer
\makeatletter
\def\ps@pprintTitle{%
 \let\@oddhead\@empty
 \let\@evenhead\@empty
 \def\@oddfoot{\it \hfill\today}%
 \let\@evenfoot\@oddfoot}
\makeatother

% A modified page layout
\textwidth 6.75in
\oddsidemargin -0.15in
\evensidemargin -0.15in
\textheight 9in
\topmargin -0.5in
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Do we need detailed demographic data to forecast plant population responses to climate change?},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

\begin{document}
\begin{frontmatter}

  \title{Do we need detailed demographic data to forecast plant population
responses to climate change?}
    \author[usu]{Andrew T. Tredennick\corref{c1}}
   \ead{atredenn(at)gmail.com} 
   \cortext[c1]{Corresponding author}
    \author[usu]{Peter B. Adler}

      \address[usu]{Department of Wildland Resources and the Ecology Center, Utah State
University, Logan, UT 84322, USA}

  \begin{abstract}
  Forecasts of populations under future climate change are constrained by
  the spatial and temporal scales at which we can gather data. For plants,
  ecologists often require individual level data to build population
  models; data that is difficult and expensive to collect at spatial
  scales greater than one square meter. Models built using such data are
  difficult to scale-up to spatial scales relevant to land management
  decisions and policy intervention. A possible route forward are
  density-structured models based on coarse-scale census data, which are
  much easier to collect over large spatial extents. However, to use
  coarse-scale data in a climate change context requires the assumption
  that the climate signals in population dynamics identified using
  individual-level data can also be identified when individual-level data
  are averaged over. Here we use a longterm dataset to model population
  dynamics using individual-level and plot-level data.
  \end{abstract}
   \begin{keyword} ecological forecasts \sep population models \sep integral projection model \sep climate change \sep grassland \sep \end{keyword}
 \end{frontmatter}

\section{Introduction}\label{introduction}

Producing skillfull forecasts at spatial scales relevant to global
environmental changes that play out at the landscape and regional scale
has proven challenging for population ecologists. The major hurdle is
that most population models, at least for plant species, are built using
data from small, localized plots because parameterizing vital rate
regressions requires tracking the fates of individuals. These models are
difficult to scale up from the micro to meso-scales because the fitted
parameters do not fully represent the spatial variation present at
spatial scales beyond that at which the data are collected.

@@

\section{Materials and Methods}\label{materials-and-methods}

\subsubsection{Study site and data}\label{study-site-and-data}

Our demographic data comes from the Fort Keogh Livestock and Range
Research Laboratory in eastern Montana's northern mixed prairie near
Miles City, Montana, USA (46 deg. 19' N, 105 deg 48' W). The dataset is
freely available on Ecological Archives (CITE), and interested readers
should refer to the metadata therin for a complete description. The site
is about 800 m above sea level and mean annual precipitation (1878-2009)
is 334 mm, with most annual precipitaion falling from April through
September (76). The site is grass dominated and, for the purposes of our
study, we focus on the four most abundant graminoid species:
\emph{Bouteloua gracilis}, \emph{Hesperostipa comata},
\emph{Pascopyrum smithii}, and \emph{Poa secunda}.

@@

\subsubsection{Stastical models of vital
rates}\label{stastical-models-of-vital-rates}

The first step in building our single-species population models was to
fit statistical models of vital rates at both levels of inference:
individual-level and quadrat-level. Here we describe the general
statistical approach at the individual-level, but the same approach
applies at the quadrat-level. We modeled survival and growth as
functions of genet size (quadrat cover in the quadrat-level models) and
climate covariates (described in more detail below). We maintained a
consistent random effects structure for both models that included three
terms: (1) a random year effect on the intercept, (2) a random year
effect on the coefficient for plant cover (either individual or quadrat
level), and (3) a random effect of group (see Data set description) on
the intercept. Vital rates for each species are modeled separately.

@@

where $x$ is the log of genet size (quadrat areal cover), $\beta_{0,t}$
is a year-specific intercept, $\beta_{s,t}$ is the year-specific slope
parameter for size, $\beta_{Q}$ is the random effect of quadrat group
location, $\theta$ is a matrix of $i$ climate effects over $t$ years,
$\beta_{c,i}$ is the fixed parameter for the effect of the $i$th climate
covariate, and $\varepsilon_{t}$ is the error term. At the quadrat level
we did not fit random year effects for survival because those parameters
did not reach convergence (see ``Computing'').

@@

\subsubsection{Including climate covariates (and avoiding model
selection)}\label{including-climate-covariates-and-avoiding-model-selection}

For both the individual-level IPM (IPM) and the quadrat-based IBM (QBM)
we included four climate covariates in each vital rate regression: fall
through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and
ppt2, respectively) and mean spring temperature at \emph{t}-1 and
\emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is
the observation year. We included climate covariates as additive
effects. We did not consider interactions among climate covariates or
among climate covariates and plant size or proportional cover. Climate
effects ($\beta_{c}$s) were modeled hierarchically in each vital rate
regression so that each species specific climate effect is drawn from an
interspecific climate effect distribution. This was necessary to achieve
convergence at the quadrat level and we used the same approach at the
individual level for uniformity.

\subsubsection{Model fitting}\label{model-fitting}

All of our analyses (model fitting and simulating) were conducted in
Program R. We used the MCMC sampler in JAGS to estimate the posterior
distributions of model parameters. We obtained posterior distributions
for all model parameters from three parallel MCMC chains run for 50,000
iterations, after a 50,000 iteration burn in. We assessed convergence
visually and using the Gelman diagnostic in the R package `coda.' Scale
reduction factors for all parameters were less than 1.02, indicating
convergence. For the purposes of introducing stochasticity in our
population models, we saved the final 1,000 iterations from each chain
for all parameters to be used as randomly drawn values during population
simulation.

\subsubsection{Population models}\label{population-models}

With the posterior distribution of the vital rate statistical models in
hand, it is straightforward to simulate the population models.

\section{Results}\label{results}

We assessed the statistical importance of the climate covariates
included the final vital rate regressions by comparing the residual
deviance of models with climate covariates and temporal random effects,
climate covariates only, and temporal random effects only. When a model
includes climate covariates, this comparison shows the relative
contribution of the climate covariates in explaining the total
interannual variability (Adler et al. 2012).

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_1.pdf}
\caption{The proportion of interannual variability in vital rates
explained by the climate covariates. The contribution for growth is
defined as: (Climate model - Constant Model)/(Full model - Constant
model). The contribution for survival and colonization, where we could
not estimate a full model with year random effects at the quadrat level,
is defined as: (Constant Model - Climate Model)/Constant Model.}
\end{figure}

\end{document}
